---
title: "Predicting Barbell Exercise Quality"
date: "7 October 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    code_folding: hide
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.8)
library(ranger)
library(rpart)
set.seed(2026)
pct <- function(x) sprintf("%.2f%%", 100 * x)
```

## Objective

This analysis predicts `classe`, the manner in which a participant performed a
barbell lift, from belt, arm, forearm, and dumbbell sensor measurements. Class A
denotes correct execution and B-E denote four prescribed mistakes. The goal is
to compare a simple classifier with a random forest, estimate performance on
unseen recording windows, and predict the 20 unlabeled course cases.

## Data and validation design

The course provides a labeled training CSV and a separate 20-row testing CSV.
The testing file has no observed `classe` and is not used for model selection or
error estimation. The source is the Weight Lifting Exercise Dataset of Velloso
and colleagues, cited below.

```{r data}
base_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/"
data_dir <- Sys.getenv("PML_DATA_DIR", "data")
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)
read_course_data <- function(filename) {
  path <- file.path(data_dir, filename)
  if (!file.exists(path)) {
    download.file(paste0(base_url, filename), path, mode = "wb",
                  quiet = TRUE)
  }
  read.csv(path, na.strings = c("NA", "#DIV/0!", ""),
           stringsAsFactors = FALSE)
}
training_raw <- read_course_data("pml-training.csv")
testing_raw <- read_course_data("pml-testing.csv")
training_raw$classe <- factor(training_raw$classe, levels = LETTERS[1:5])
stopifnot(nrow(training_raw) == 19622, nrow(testing_raw) == 20)

group <- paste(training_raw$user_name, training_raw$num_window, sep = ":")
first <- !duplicated(group)
group_table <- data.frame(group = group[first],
                          classe = training_raw$classe[first])
stopifnot(all(vapply(split(training_raw$classe, group),
                     function(x) length(unique(x)), integer(1)) == 1))

build_groups <- unlist(lapply(split(group_table$group, group_table$classe),
                             function(x) sample(x, floor(0.75 * length(x)))),
                       use.names = FALSE)
in_build <- group %in% build_groups
stopifnot(length(intersect(unique(group[in_build]),
                           unique(group[!in_build]))) == 0)

candidate_names <- names(training_raw)[8:159]
usable <- vapply(training_raw[in_build, candidate_names], function(x) {
  is.numeric(x) && !anyNA(x) && length(unique(x)) > 1
}, logical(1))
predictor_names <- candidate_names[usable]
model_data <- training_raw[c(predictor_names, "classe")]
build <- model_data[in_build, ]
validation <- model_data[!in_build, ]
quiz <- testing_raw[predictor_names]
stopifnot(length(predictor_names) == 52, !anyNA(validation), !anyNA(quiz))

knitr::kable(data.frame(
  Class = levels(model_data$classe),
  Build = as.integer(table(build$classe)),
  Validation = as.integer(table(validation$classe))),
  caption = "Class counts after the grouped split")
```

There are `r nrow(training_raw)` labeled rows, `r ncol(training_raw)` columns,
six participants, and `r nrow(group_table)` participant/window groups. Repeated
readings in a recording window are related. A random row split could therefore
give an optimistic estimate. Approximately 75% of windows within each class
were assigned to model building and the remainder to an untouched validation
set. This produced `r nrow(build)` build rows in
`r length(unique(group[in_build]))` windows and `r nrow(validation)` validation
rows in `r length(unique(group[!in_build]))` windows.

The first seven fields are identifiers, participant names, timestamps, and
window descriptors. They define groups but are excluded as predictors. Feature
screening used only the build partition. The `r length(predictor_names)` retained
measurements are numeric, complete, and nonconstant. Most excluded sensor
summary fields are almost entirely missing. Selected measurements are also
complete in validation and the quiz file, so no imputation is needed. Tree
models do not require scaling. No feature selection used validation outcomes.

## Models and grouped cross-validation

A single classification tree provides an interpretable baseline. A random
forest can capture nonlinear boundaries and interactions while averaging
many trees to reduce variance. Five-fold cross-validation was performed within
the build partition. Windows, stratified by class, were assigned to folds as
whole groups. All candidate models used exactly the same folds.

For the forest, the number of candidate variables per split (`mtry`) was tested
at 2, 7, and 14. Seven is near the square root of 52. Each tuning fit used 300
trees, Gini splitting, and a minimum node size of one. The baseline used
`rpart` with complexity parameter 0.01. Mean fold accuracy selected the forest
setting. The validation set did not participate in this decision.

```{r cross-validation}
build_group_table <- group_table[group_table$group %in% build_groups, ]
build_group_table$fold <- 0L
for (class in levels(model_data$classe)) {
  idx <- which(build_group_table$classe == class)
  build_group_table$fold[idx] <- sample(rep(1:5, length.out = length(idx)))
}
row_fold <- build_group_table$fold[
  match(group[in_build], build_group_table$group)]
cv_results <- data.frame()
for (fold in 1:5) {
  fit_rows <- row_fold != fold
  test_rows <- !fit_rows
  tree <- rpart(classe ~ ., data = build[fit_rows, ], method = "class",
                control = rpart.control(cp = 0.01, xval = 0))
  tree_prediction <- predict(tree, build[test_rows, ], type = "class")
  cv_results <- rbind(cv_results, data.frame(
    Model = "Single tree", mtry = NA_integer_, Fold = fold,
    Accuracy = mean(tree_prediction == build$classe[test_rows])))
  for (m in c(2, 7, 14)) {
    forest <- ranger(classe ~ ., data = build[fit_rows, ],
                      num.trees = 300, mtry = m, min.node.size = 1,
                      splitrule = "gini", seed = 2026 + 100 * fold + m,
                      num.threads = 2)
    prediction <- predict(forest, build[test_rows, ])$predictions
    cv_results <- rbind(cv_results, data.frame(
      Model = "Random forest", mtry = m, Fold = fold,
      Accuracy = mean(prediction == build$classe[test_rows])))
  }
}
settings <- c("Single tree", "RF: mtry = 2", "RF: mtry = 7", "RF: mtry = 14")
cv_results$Setting <- ifelse(cv_results$Model == "Single tree", "Single tree",
                             paste0("RF: mtry = ", cv_results$mtry))
cv_summary <- do.call(rbind, lapply(settings, function(setting) {
  a <- cv_results$Accuracy[cv_results$Setting == setting]
  data.frame(Setting = setting, Mean_accuracy = mean(a),
             Fold_SD = sd(a), Mean_error = 1 - mean(a))
}))
best <- which.max(cv_summary$Mean_accuracy[-1])
best_mtry <- c(2, 7, 14)[best]
best_cv <- cv_summary$Mean_accuracy[best + 1]
cv_display <- cv_summary
cv_display[-1] <- lapply(cv_display[-1], function(x) sprintf("%.2f%%", 100 * x))
knitr::kable(cv_display, col.names = c("Model", "Mean accuracy",
                                      "Fold SD", "Mean error"))
```

The selected `mtry` is `r best_mtry`, with mean cross-validation accuracy
`r pct(best_cv)` and estimated error `r pct(1 - best_cv)`. The table distinguishes
fold-to-fold standard deviation from a confidence interval. Tuning is limited
to a small grid to keep the analysis practical and reproducible.

## Untouched validation performance

After choosing `mtry`, the forest was trained on the whole build partition
with 500 trees. This increases ensemble size without using validation labels
to tune the model. Accuracy and the confusion matrix were calculated once on
the reserved windows.

```{r validation}
validation_model <- ranger(classe ~ ., data = build,
                            num.trees = 500, mtry = best_mtry,
                            min.node.size = 1, splitrule = "gini",
                            importance = "permutation", seed = 2026,
                            num.threads = 2)
validation_prediction <- predict(validation_model, validation)$predictions
cm <- table(Predicted = validation_prediction, Observed = validation$classe)
correct <- sum(diag(cm))
accuracy <- correct / sum(cm)
errors <- sum(cm) - correct
recall <- diag(cm) / colSums(cm)
knitr::kable(cm, caption = "Validation confusion matrix: rows predicted, columns observed")
knitr::kable(data.frame(Class = names(recall), Recall = sprintf("%.2f%%", 100 * recall)))

validation_group <- group[!in_build]
window_correct <- tapply(validation_prediction == validation$classe,
                         validation_group, sum)
window_n <- table(validation_group)[names(window_correct)]
set.seed(2027)
bootstrap_accuracy <- replicate(2000, {
  idx <- sample.int(length(window_correct), replace = TRUE)
  sum(window_correct[idx]) / sum(window_n[idx])
})
accuracy_ci <- quantile(bootstrap_accuracy, c(0.025, 0.975))
```

The model correctly classified `r correct` of `r sum(cm)` validation readings,
with `r errors` errors. **Validation accuracy is `r pct(accuracy)` and estimated
out-of-sample error is `r pct(1 - accuracy)`.** A percentile bootstrap resampling
the validation windows 2,000 times gives an approximate 95% accuracy interval
of `r pct(accuracy_ci[1])` to `r pct(accuracy_ci[2])`. Resampling whole windows
preserves within-window dependence. The interval reflects validation sampling
uncertainty conditional on the fitted model, not all model-selection uncertainty.

```{r importance-figure, echo=FALSE, fig.cap="Top 12 measurements by permutation importance in the build-partition forest. Importance describes predictive contribution, not causality."}
importance <- sort(validation_model$variable.importance, decreasing = TRUE)
top <- head(importance, 12)
par(mar = c(4, 11, 2, 1), las = 1)
barplot(rev(top), horiz = TRUE, col = "#427c8e", border = NA,
         xlab = "Permutation importance", cex.names = 0.85)
```

The highest-ranked measurement is `r names(importance)[1]`. Permutation
importance can be affected by correlated sensors, so the ranking is descriptive.
The single-tree comparison and the holdout assessment provide separate evidence
for choosing the forest.

## Predictions for the 20 course cases

Only after validation was complete, a final 500-tree forest with the selected
settings was fitted to all labeled rows. The same 52 predictors were retained.
The table is sorted by `problem_id`, rather than relying on file row order.

```{r final-predictions}
final_model <- ranger(classe ~ ., data = model_data,
                       num.trees = 500, mtry = best_mtry,
                       min.node.size = 1, splitrule = "gini",
                       seed = 2028, num.threads = 2)
quiz_prediction <- predict(final_model, quiz)$predictions
answers <- data.frame(problem_id = testing_raw$problem_id,
                      prediction = as.character(quiz_prediction))
answers <- answers[order(answers$problem_id), ]
stopifnot(identical(answers$problem_id, 1:20),
           all(answers$prediction %in% LETTERS[1:5]))
knitr::kable(answers)
writeLines(paste(answers$problem_id, answers$prediction, sep = ": "),
            "predictions.txt")
```

Predictions in problem order:
**`r paste(answers$prediction, collapse = ", ")`**.

## Conclusion and limitations

The random forest predicts exercise class with an estimated error of
`r pct(1 - accuracy)` on reserved recording windows. This is an estimate for
similar measurements under the study protocol, not a guarantee for each quiz
case. No labels are available to measure accuracy on the 20 course cases.

Window grouping reduces one source of leakage, but the build and validation
partitions still contain the same six participants. Nearby windows may also
be related. Performance for new people, recording sessions, or devices may be
lower. A larger study with participant- or session-held-out validation would
be needed for a deployment claim.

## Reproducibility and sources

The report was compiled with `knitr` and `rmarkdown`. To reproduce it, install
`rmarkdown`, `knitr`, `ranger`, and `rpart`, then run
`rmarkdown::render("Prediction_Assignment_Writeup.Rmd")` from the project
directory. Missing data files are downloaded from the course links. Fixed seeds
specify splitting and model fitting. The HTML includes evaluated results and
one figure, so peer reviewers do not need to run R.

Velloso, E., Bulling, A., Gellersen, H., Ugulino, W., and Fuks, H. (2013).
*Qualitative Activity Recognition of Weight Lifting Exercises*. Proceedings
of the 4th Augmented Human International Conference, pp. 116-123.
[DOI: 10.1145/2459236.2459256](https://doi.org/10.1145/2459236.2459256).
[Archived dataset description](https://web.archive.org/web/20161224072740/http://groupware.les.inf.puc-rio.br/har).

Data: [training CSV](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv)
and [testing CSV](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv).

```{r session, echo=FALSE}
sessionInfo()
```
